Python to Assembly Compilation

### 2011-12-14 paper/paper.tex
- Changes up through 3.3.1 - Expressions

The \texttt{pymem\_new} routine works by allocating slightly more memory than is requested by the user.  This extra memory is used to store a pointer to a linked list node that contains the details of the allocation, including the type of object being created, the requested size, the pointer returned by \texttt{malloc} and the time of allocation.  The caller is then returned the portion of memory immediately after this pointer.  Subsequent calls to \texttt{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as de-allocated, as well as the de-allocation time.  The memory can then be freed independently of the linked list containing the statistics. 
\begin{figure}
\begin{center}
\includegraphics[scale=0.48]{compiler_flow.pdf}
\end{center}
\caption{Compiler Architecture}
\label{fig-comparch}
\end{figure}

\subsection{Compiler Flow}
Refer to Figure~\ref{fig-comparch} for the compiler data flow.  Each stage operates on an abstract syntax tree, the relevant new stages are the garbage collection flattening stage and the reference count AST transformation stage.

\subsection{Syntax Tree Modifications}

\subsubsection{New compiler stages}
We added two new stages to our reference compiler.  The first stage is a modified flattening stage, which produces a flattened AST to aid in the reference determination.  The next stage is the actual reference determination and AST transformation, where the compiler determines what variables will be objects and when assignments happen and adds appropriate calls to increment and decrement the reference counter of the object as needed.
\par
Our initial assumption was that we could easily add the increment and decrement reference counter operations in the reference compilers flatten stage, however because this stage is after type explication and there are variables created which do not correspond to a type in the language the pre-flatten stage was necessary.  This stage converts all expressions in to assignments to temporary variables.  At this stage we know every expression is operating on types.

\subsubsection{Analysis on when to decrement an object reference}
The next new stage is the reference counting AST transformation.  In this stage we take the flattened tree and look for assignments which are setting list elements, dictionary values, class attributes, object attributes or variables.  The compiler determines what variables are created in the current scope and injects code at the beginning of that scope to set the value initially.  We decided on this approach to simplify determining when an object is first assigned and this allows us to simply decrement the reference counter to the object before an assign, regardless of the variables state.

@@

\subsection{Interesting Problems}
We encountered a number of interesting questions during this implementation and summarize them below.

@@

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

>>>>

The \texttt{pymem\_new} routine works by allocating slightly more memory than is requested by the user.  This extra memory is used to store a pointer to a linked list node that contains the details of the allocation, including the type of object being created, the requested size, the pointer returned by \texttt{malloc} and the time of allocation.  The caller is then returned the portion of memory immediately after this pointer.  Subsequent calls to \texttt{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as de-allocated, as well as the de-allocation time.  The memory can then be freed independently of the linked list containing the statistics.

\subsection{Compiler}

Figure~\ref{fig-comparch} shows the compiler data flow after modifying our compiler for reference counting.  Each stage operates on an abstract syntax tree (AST).  The relevant new stages are the pre-explicate flatten stage, and the reference counting AST transformation stage.

\begin{figure}
\begin{center}
\includegraphics[scale=0.48]{compiler_flow.pdf}
\end{center}
\caption{Compiler Architecture}
\label{fig-comparch}
\end{figure}

\subsubsection{Reference Counting AST Transformation}

Recall that in order to properly maintain reference counts, we need to increment the reference count for an object when a reference is created for it, and decrement the reference count when a reference is destroyed.  References are created by assigning a memory location (which contains an object) to a variable.  References are destroyed when a variable is changed to refer to a different object, or when a variable ceases to exist when a program reaches the end of a given scope.

The most natural way to implement this in our compiler, is to transform our input programs AST such that any assignment to a variable is preceded by a decrement to the reference count for the object the variable refers to, and is succeeded by an increment to the reference count for the newly assigned object.  This is the approach we took, but there are some subtleties with respect to where this operation is placed within the compiler.  Our initial hypothesis was that we could easily add the increment and decrement reference counter operations in the existing ``Flatten AST'' stage.  We quickly realized that because this stage is after type explication, there exist variables which are not \textit{tagged} with a type.  Since we only want to decrement reference counts for \textit{tagged} types\footnote{Actually, we only want to modify reference counts for a subset of tagged types that only includes ``big'' types. However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \texttt{int} and \texttt{boolean}.}, the only option is to put this transformation stage before type explication.

\subsection{Challenges}
We encountered a number of interesting challenges related to our design choices in modifying the compiler and summarize them below.

\subsubsection{Expressions}
Prior to our explicate phase, complex expressions are not flattened.  This leads to problems where memory is leaked, since no assignments ever occur to temporary objects.  Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \texttt{a}.  The assignment to the variable \texttt{a} results in a corresponding increment reference count on the result of the list addition.  However, the temporary lists, \texttt{[1]} and \texttt{[2]} are leaked if we don't take any additional steps.  To overcome this problem, we introduce a intermediate flattening stage, prior to our reference counting AST transformation.   This results in the corresponding code given in Listing~\ref{lst:flattened}.

\begin{figure}[h!]
\begin{lstlisting}[
language=Python,
frame=topline,
caption={A simple input program that leaks memory if not flattened.},
captionpos=b,
label=lst:simpleleak
]
a = [1] + [2]
\end{lstlisting}
\end{figure}

\begin{figure}[h!]
\begin{lstlisting}[
language=Python,
frame=topline,
caption={The flattened version of listing \ref{lst:simpleleak}.},
captionpos=b,
label=lst:flattened
]
tmp0 = [1]
tmp1 = [2]
tmp2 = tmp0 + tmp1
\end{lstlisting}
\end{figure}

@@

@@

%\appendix
%\section{Appendix Title}
% This is the text of the appendix, if you need one.
